---
title: |
        <span style='font-weight: bold;'>Rain in Australia</span>
        
        Forecasting the amount of rainfall, in mm, using the Box-Jenkins Methodology
author:
  - name: Carolina Almeida 20221855

  - name: João Capitão 20221863

  - name: Teresa Simão 20221873


affiliation:
 -num: 1
 address: Bachelor in Data Science, NOVA Information Management School
 
column_numbers: 3
 
author_textsize: '60pt'
title_textsize: '60pt'

output: 
  posterdown::posterdown_html:
    self_contained: true
    css: Styles.css
    
bibliography: Packages.bib
knit: pagedown::chrome_print

primary_colour: '#7e96ab'
secondary_colour: '#39657b'
accent_colour: '#1c3953'

#columnline_col: '#7e96ab'
#columnline_width: '5mm'
columnline_style: 'none'

title_textcol: '#000000'
author_textcol: '#000000'
affiliation_textcol: '#000000'

body_textsize: '40px'

font_family: 'New Century Schoolbook'
titletext_fontfamily: 'New Century Schoolbook'

logoleft_name: C:/Users/Carolina/OneDrive - NOVAIMS/Desktop/Second Year - Second Semester/Forecasting Methods/Practical Classes/Group Project/Project/Logo.png

editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# <span style='color: black;'>Introduction</span>

From tropical downpours to arid landscapes, understanding and predicting rain patterns is essential for managing resources and ensuring resilience in the face of changing climates. In Australia, the significance of this task transcends various sectors, spanning agriculture, water resource management by optimizing water usage, and public safety by helping mitigate the impacts of floods and droughts.
This study contributes to the ongoing efforts in climate science to improve predictive capabilities and support informed decision-making in weather-dependent sectors.

```{r include = FALSE}

library(fpp3)
library(lubridate)
library(urca)
library(tidyverse)
library(tsibble)
library(forecast)  
library(dplyr)

rain_in_australia <- read.csv('data_group_L.csv')

View(rain_in_australia)

# Format the 'Date' column by replacing the hyphens by slashes
rain_in_australia <- rain_in_australia %>% mutate(Date = gsub("-", "/", Date))

# Transform the dataset into a tsibble
rain_in_australia_ts <- rain_in_australia %>%
  mutate(Month = yearmonth(Date)) %>%
  group_by(Month) %>% 
  summarise(Rainfall = mean(Rainfall, na.rm = TRUE)) %>% 
  select(Month, Rainfall) %>%
  as_tsibble(index = Month)

View(rain_in_australia_ts)

# Note: Rainfall is the amount of rainfall recorded for the day in mm

# Handling missing values

# Some periods don't have data
# Feeling the gaps with NaNs
rain_in_australia_ts <- fill_gaps(rain_in_australia_ts)

# Replacing any missing values in the 'Rainfall' column with the mean value.
mean_value <- mean(rain_in_australia_ts$Rainfall, na.rm = TRUE)
rain_in_australia_ts$Rainfall <- ifelse(is.na(rain_in_australia_ts$Rainfall), mean_value, rain_in_australia_ts$Rainfall)

# We don't have missing values anymore
missing_values_per_column <- sapply(rain_in_australia_ts, function(x) sum(is.na(x)))
print(missing_values_per_column)

# Plots
rain_in_australia_ts %>% gg_tsdisplay(Rainfall, plot_type = 'partial')

rain_in_australia_ts %>% gg_season(Rainfall)

```

```{r out.width = '75%', fig.align = 'center', fig.cap = "Daily amount of rainfall (mm)", fig.height = 5 }
rain_in_australia_ts %>% autoplot(Rainfall,  color = "#7e96ab") + 
  theme(panel.grid.major = element_line(color = "#eeeeee"),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white"))
```

```{r include = FALSE}

# Non - stationary series
# We do not have a trend

rain_in_australia_ts %>% gg_subseries(Rainfall)

# Training and Testing sets

# Deciding the proportion we will have for training and testing sets
start_date <- as.Date("2007-11-01")
end_date <- as.Date("2017-06-25")
threshold_date <- as.Date("2015-01-01")
total_days <- as.numeric(end_date - start_date)
train_days <- as.numeric(threshold_date - start_date) 
test_days <- as.numeric(end_date - threshold_date) 

cat(round((train_days/total_days) * 100, 2), '% of the dataset will be training data', round((test_days/total_days) * 100, 2), '% of the dataset will be test data')

# Splitting the data into train and test datasets
rain_in_australia_ts_train <- rain_in_australia_ts %>% filter(year(Month) < 2015)
rain_in_australia_ts_test <- rain_in_australia_ts %>% filter(year(Month) >= 2015)

# Variance analysis

rain_in_australia_ts_train %>%
  ggplot(aes(y = Rainfall, x = Month)) +
  geom_line(color = 'black') + labs(title = "Daily amount of rainfall (mm)", x = 'Months [1M]', y = 'Rainfall') + 
  theme_minimal()

# The variance is not stable over time

# Check the log transformation
rain_in_australia_ts_train %>%
  ggplot(aes(x = Month, y = log(Rainfall))) +
  geom_line(color = 'black') + labs(title = "Log-transformed Rainfall", x = 'Months [1M]', y = 'Log(Rainfall) (log(mm))') + 
  theme_minimal()

# After doing this we notice a very big negative peak, which indicates the presence of zero or near-zero values.
# After looking at the dataset, we notice that 2008 May has a value very close to zero. Meaning that there were days with very low rainfall amounts

# Therefore we will apply Box-Cox transformation

# Estimating the lambda
rainfall_ts <- ts(rain_in_australia_ts_train$Rainfall, start = c(2007, 11), frequency = 12)
lambda_guerrero <- guerrero(rainfall_ts)

rain_in_australia_ts_train$AdjustedRainfall <- BoxCox(rain_in_australia_ts_train$Rainfall, lambda_guerrero)

rain_in_australia_ts_train %>%
  ggplot(aes(x = Month, y = AdjustedRainfall)) +
  geom_line(color = 'black') + labs(title = "Adjusted Rainfall", x = 'Months [1M]', y = 'Adjusted Rainfall') + 
  theme_minimal()

# Testing if it is Stationary - Augmented Dickey-Fuller (ADF) test
summary(ur.df(rain_in_australia_ts_train$AdjustedRainfall, type = "none", lags = 12))

# The test statistic (-0.2327) is greater than all the critical values (-2.6, -1.95, -1.61) at the 1%, 5%, and 10% significance levels.
# This suggests that the test statistic is not significant at any conventional level of significance.
# Therefore, we fail to reject the null hypothesis of a unit root, indicating that the series is likely non-stationary.

# So, we will take the differences
# First, the seasonal difference

rain_in_australia_ts_train <- rain_in_australia_ts_train %>%
  mutate("seasonal_diff_val" = difference(AdjustedRainfall, lag = 12)) %>% na.omit()

rain_in_australia_ts_train %>%
  ggplot(aes(y = seasonal_diff_val, x = Month))+
  geom_line(color = 'black') + labs(title = "Adjusted Rainfall- Seasonal Differenced", x = 'Months [1M]', y = 'Adjusted Rainfall') + theme_minimal()

# Note: For the first 12 months of our original time series, we won't have corresponding seasonal difference values because there are no observations 12 months before those.

rain_in_australia_ts_train %>%
  gg_tsdisplay(seasonal_diff_val, plot_type = 'partial') + labs(title = "Adjusted Rainfall- Seasonal Differenced", x = 'Months [1M]', y = 'Adjusted Rainfall')

# Testing if the seasonally differenced series is stationary - ADF test
summary(ur.df(rain_in_australia_ts_train$seasonal_diff_val, type = 'none', lags = 12))

# The test statistic (-2.1818 ) is smaller than the critical value at 5% significance level (-1.95). So, we reject H0 at 5% significance level.
# There is statistical evidence that the seasonally differenced series is stationary
```

# <span style='color: black;'>Methodology</span>

Initially, the gaps and missing values were addressed. The latter was managed through a systematic approach: replacing missing values with the mean (since it maintained the dataset's central tendency and preserved the overall data structure). Following this, the dataset was divided into two subsets: a validation set encompassing years from 2008 to 2014, and a test set containing data from 2015 to 2017. Our goal was to achieve a balanced split of 75% for training and 25% for testing, a target we successfully met.To stabilize the variance of our training time series, we applied a **Box-Cox transformation**. Recognizing that the data still lacked stationarity, we then implemented a seasonal differencing technique. The resulting series was confirmed to be stationary through a **unit root test**.

```{r out.width = '75%', fig.align = 'center', fig.cap = "Daily amount of rainfall (mm) after differentiation; ACF Plot; PACF Plot", fig.height = 5 }

rain_in_australia_ts_train %>% 
  gg_tsdisplay(seasonal_diff_val, plot_type = 'partial', lag_max = 36) + 
  labs(title = "Transformed and Seasonally Differenced Rainfall", x = 'Months [1M]', y = 'Adjusted Rainfall')

```

Based on the autocorrelation and partial autocorrelation plots obtained, several ARIMA models were selected and evaluated, using the following criteria:

- d = 0 since differencing was not applied to achieve stationarity
- D = 1 since seasonal differencing was applied

* p is typically set to 2 due to a spike at lag 2 in the PACF plot
* q is sometimes set to 2 due to a spike at lag 2 in the ACF plot

+ P is generally set to 0 due to the absence of noticeable spikes at lags 12, 24, and 36 in the PACF plot
+ Q is usually set to 1 due to a spike at lag 12 in the ACF.

Additionally, the **Ljung-Box test** was applied to these models, and none showed statistical evidence of autocorrelation in the residuals.

# <span style='color: black;'>Final Results</span>

```{r include = FALSE}

# Model Identification - Selecting the candidate models

fit_1 <- rain_in_australia_ts_train %>%
  model(
    sarima002011 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(0,0,2) + PDQ(0,1,1)),
    sarima200011 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2,0,0) + PDQ(0,1,1)),
    sarima202011 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2,0,2) + PDQ(0,1,1)),
    sarima002010 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(0,0,2) + PDQ(0,1,0)),
    sarima200010 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2,0,0) + PDQ(0,1,0)),
    sarima200111 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2,0,0) + PDQ(1,1,1)),
    sarima200012 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2, 0, 0) + PDQ(0,1,2)),
    auto_ARIMA = ARIMA(box_cox(Rainfall, lambda_guerrero)))

# ARIMA (p,d,q) (P,D,Q):

# d = 0 since differencing was not applied to achieve stationarity
# D = 1 since seasonal differencing was applied

# p is usually set to 2 since we have a spike in lag 2 in the PACF plot
# q is sometimes set to 2 due to a spike in lag 2 in the ACF plot

# P is usually set to 0 due to no noticeable spikes at lags 12, 24, 36 in the PACF plot
# Q is usually set to 1 due to a spike at lag 12 in ACF plot

fit_1 %>% glance() 

# Residuals, ljung_box test and box_pierce test

# sarima002011
fit_1 %>% select(sarima002011) %>% report()

fit_1 %>%
  select(sarima002011) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima002011", x = 'Months [1M]')

fit_1 %>%
  select(sarima002011) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima002011) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima200011
fit_1 %>% select(sarima200011) %>% report()

fit_1 %>%
  select(sarima200011) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima200011", x = 'Months [1M]')

fit_1 %>%
  select(sarima200011) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima200011) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima202011
fit_1 %>% select(sarima202011) %>% report()

fit_1 %>%
  select(sarima202011) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima202011", x = 'Months [1M]')

fit_1 %>%
  select(sarima202011) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima202011) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima002010
fit_1 %>% select(sarima002010) %>% report()

fit_1 %>%
  select(sarima002010) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima002010", x = 'Months [1M]')

fit_1 %>%
  select(sarima002010) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima002010) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima200010
fit_1 %>% select(sarima200010) %>% report()

fit_1 %>%
  select(sarima200010) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima200010", x = 'Months [1M]')

fit_1 %>%
  select(sarima200010) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima200010) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima200111
fit_1 %>% select(sarima200111) %>% report()

fit_1 %>%
  select(sarima200111) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima200111", x = 'Months [1M]')

fit_1 %>%
  select(sarima200111) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima200111) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# sarima200012
fit_1 %>% select(sarima200012) %>% report()

fit_1 %>%
  select(sarima200012) %>%
  gg_tsresiduals() + labs(title = "Residuals of sarima200012", x = 'Months [1M]')

fit_1 %>%
  select(sarima200012) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(sarima200012) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# auto_ARIMA
fit_1 %>% select(auto_ARIMA) %>% report()

fit_1 %>%
  select(auto_ARIMA) %>%
  gg_tsresiduals() + labs(title = "Residuals of auto_ARIMA", x = 'Months [1M]')

fit_1 %>%
  select(auto_ARIMA) %>%
  augment() %>%
  features(.innov, ljung_box, lag = 36) 

fit_1 %>%
  select(auto_ARIMA) %>%
  augment() %>%
  features(.innov, box_pierce, lag = 36)

# The p-values obtained from these tests are all greater than 5%
# We fail to reject the null hypothesis - There is no statistical evidence for the presence of autocorrelation in the residuals.
# We can proceed to the final step: forecasting

fit_1 %>% forecast(h = '24 months') %>%
  accuracy(rain_in_australia_ts_test)

# auto_ARIMA and sarima200111 have the lowest error, so, they will be compared with Seasonal Naive

final_fit <- rain_in_australia_ts_train %>%   
  model(snaive = SNAIVE(box_cox(Rainfall, lambda_guerrero)),
  sarima200111 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2,0,0) + PDQ(0,1,1)),
  sarima200012 = ARIMA(box_cox(Rainfall, lambda_guerrero) ~ pdq(2, 0, 0) + PDQ(0,1,2)))

final_fit %>%
  forecast(h = '24 months') %>%
  accuracy(rain_in_australia_ts_test)

```

After completing the evaluation and analyzing the errors, illustrated bellow, it was concluded that SARIMA(2,0,0)(1,1,1) and SARIMA(2,0,0)(0,1,2) were the best models.

```{r out.width = '75%', fig.align = 'center', fig.cap = "Errors for each model", fig.height = 5 }

table <- fit_1 %>% forecast(h = '24 months') %>%
  accuracy(rain_in_australia_ts_test)

knitr::kable(table[, c(".model", ".type", "ME", "RMSE", "MAE", "MPE", "MAPE")], align = "c")

```

It can be inferred that both of these models effectively captured the historical patterns of rainfall in Australia over recent years.

```{r out.width = '75%', fig.align='center', fig.cap = "Forecast of amount of rain in Australia - Seasonal Naive; SARIMA(2,0,0)(1,1,1); sarima(2,0,0)(0,1,2)", fig.height = 5 }

final_fit %>%
  forecast(h = '30 months') %>%
  autoplot(rain_in_australia_ts_train, level = FALSE) +
  scale_fill_manual(values = c("#4bac39","#7e96ab","#6a329f")) +
  scale_color_manual(values = c("#4bac39","#7e96ab","#6a329f")) +
  ggtitle("Forecasts of Amount of Rain in Australia") +
  xlab("Months [1M]") +
  ylab("Total Amount of Rain, in mm") +
  theme_minimal()

```

# <span style='color: black;'>Conclusion</span>

Among the models considered, SARIMA(2,0,0)(0,1,2) emerged with the lowest RMSE (0.944) and MAE (0.756), indicating superior performance in terms of these error metrics. Additionally, the ME value being very close to zero (-0.067) suggests minimal bias in the predictions. On the other hand, SARIMA(2,0,0)(1,1,1), also demonstrated commendable performance with low RMSE (0.943) and MAE (0.756) values. Also, similar to SARIMA(2,0,0)(0,1,2), its ME and ACF1 values are close to zero, indicating satisfactory model performance.

Thus, the selected model is **SARIMA(2,0,0)(1,1,1)**.

```{r include = FALSE}

# Among the models considered, SARIMA(2,0,0)(0,1,2) emerged with the lowest RMSE (0.944) and MAE (0.756), 
# indicating superior performance in terms of these error metrics. 
# Additionally, the ME value being very close to zero (-0.067) suggests minimal bias in the predictions. 
# On the other hand, SARIMA(2,0,0)(1,1,1), also demonstrated commendable performance with low RMSE (0.943) and MAE (0.756) values. 
# Also, similar to SARIMA(2,0,0)(0,1,2), its ME and ACF1 values are close to zero, indicating satisfactory model performance.

# Selected model: SARIMA(2,0,0)(1,1,1)

final_fit %>% select(sarima200111) %>% report()

```

```{r out.width = '75%', fig.align = 'center', fig.cap = "Forecast of amount of rain in Australia", fig.height = 5 }

# Forecasting
final_fit %>%
  select(sarima200111) %>%
  forecast(h = 30) %>%
  autoplot(rain_in_australia_ts_test, level = NULL, color = "#7e96ab") +
  ggtitle("Amount of Rainfall recorded in Australia, in mm") +
  labs(subtitle = "Forecasts of SARIMA2000111") +
  xlab("Months [1M]") + ylab("Rainfall") +
  theme(panel.grid.major = element_line(color = "#eeeeee"),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white"))
```

It is important to note that we analyzed the residuals of our model using a plot, and confirmed that they exhibit characteristics of white noise and follow a normal distribution.

```{r include = FALSE}
final_fit %>%
  select(sarima200111) %>%
  gg_tsresiduals() + labs(title = "Residuals of SARIMA200111 model", x = 'Months [1M]')

```

To conclude, despite the challenges posed by the dynamic nature of climate, We were able to successful attain a robust and consistent model to forecast rainfall.
Our findings underscore the importance of employing sophisticated time series analysis techniques in tackling the complexities of climate prediction. By leveraging SARIMA models, we equip stakeholders with valuable insights to inform decision-making and resource management strategies in the face of Australia's ever-changing weather patterns.

```{r include = FALSE}

# Compute the difference between the forecast and the reality

forecast_values <- sum(final_fit %>% select(sarima200111) %>% forecast(h = 30.5) %>% filter(year(Month) >= 2015) %>% pull(.mean))

real_values <- sum(rain_in_australia_ts_test %>% pull(Rainfall))

(real_values / forecast_values) * 100 # Correct values predicted

```

```{r, include  = FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```

# <span style='color: black;'>References</span>

[1] Time Series Basics and Getting Started with R

[2] Forecaster's Toolbox 

[3] Introduction to Time Series. ARMA models 

[4] Box - Jenkins methodology. Forecasting with ARIMA models. Introduction to Seasonal ARIMA models 

Data acquired from:
<https://www.kaggle.com/datasets/jsphyg/weather-dataset-rattle-package>